%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{etoolbox}
\makeatletter
\patchcmd{\maketitle}{\@copyrightspace}{}{}{}
\makeatother


\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

% \conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
% \copyrightyear{20yy}
% \copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
% \doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish,
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers,
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Open Artifacts}
% \subtitle{Subtitle Text, if any}

\authorinfo{Heather Miller}
           {EPFL}
           {heather.miller@epfl.ch}
\authorinfo{Philipp Haller}
           {KTH Royal Institute of Technology}
           {philipp.haller@a3.epfl.ch}

\maketitle


Open source software (OSS) development is more successful today than it has
ever been. It has enabled research projects such as LLVM and Apache Spark to
``escape the lab" and create thriving open-source communities with large
numbers of industrial contributors (for example, developers from over 50
companies have contributed to the development of Apache Spark). Conversely,
companies like Twitter and Netflix are developing core components of their
software stacks as popular open-source projects that are reused by numerous
other companies and organizations.

Does the PL research community not strive for some of the same goals, namely,
that our work is reused and built upon by many others? (This practice would
result in ``citations for free''!)

In this talk we argue that adopting tools and methodologies of modern OSS
development practices (that go far beyond public version control systems)
would enable much greater reusability and extensibility of published software
artifacts. Furthermore, our proposal could make researchers more productive by
significantly reducing the effort required to build upon previously published
artifacts.

In recent years, leading PL conferences (POPL, PLDI, OOPSLA, ECOOP, and
others) have introduced artifact evaluation (AE)
processes.\footnote{http://www.artifact-eval.org/} While AE processes are an
important step towards increasing the reproducibility of PL
research~\cite{Vitek14}, they are not without issues. We argue that the
current generation of AE processes (a) gives away a lot of potential to
encourage truly reusable research artifacts, and (b) tends to accommodate
poorly-developed artifacts, which, as a result, are relevant to only a small
audience; at the same time, the evaluation of such artifacts requires a
significant time investment by the AE committee (AEC).

Meanwhile, following decades of development on OSS, big for-profit companies
like Microsoft and Oracle are pressured into joining the OSS movement in order
to stay competitive. Besides offering social benefits with respect to the
developer community, the OSS community has developed tools and methodologies
whose effectiveness in terms of enabling growth and adoption is measurable.

Our own experience with the AE processes of two major PL conferences (POPL and
OOPSLA) on both sides of the reviewing process has shown us, though, that many
of the principles that, arguably, make OSS development so successful are
largely ignored when producing and evaluating artifacts, in effect making them
much less valuable than they could be, both for the academic community and
society at large.

Why do we ignore decades worth of work on reproducibility and openness that is
at the heart of a thriving industry?

Active OSS projects are often characterized by a vivid, organically evolving
environment of people being inspired by one another and building off of each
other's work. Arguably, such an environment is something we strive for in
academia, too. Unfortunately, in reality it is often hard to reuse and build
upon the software artifacts of others. Moreover, the visibility of software
artifacts remains limited if they are only available from academic web pages.
In contrast, OSS development thrives in part thanks to centralized software
directories such as Hackage and GitHub which make it easy to find or stumble
upon interesting OSS projects and packages. Social features of popular code
hosting platforms reinforce such aspects.

In contrast, it is surprisingly common for artifacts submitted in the context
of the AE processes of major conferences to be bundled as virtual machine
images. Virtual machine images reduce the effort to get artifacts to run on
machines of AEC members. At the same time, they enable and, to some extent,
even encourage researchers to avoid or neglect aspects of their artifacts that
would allow others to reuse them. To put it bluntly: by providing a virtual
machine image, not even the build of an artifact has to be repeatable! Virtual
machines are problematic for another reason: benchmark results often cannot be
reproduced faithfully due to interference with process scheduling, memory
access, and others. On the other hand, artifacts that need a carefully-tuned
environment like a new OS kernel may only be practical using virtual machines.

Why don't we follow the trends in OSS development and thereby (a) resolve some
of the above-mentioned issues, and (b) significantly increase the impact of
research results that are accompanied by ``open artifacts''?

Most modern OSS projects are hosted on social code platforms, such as GitHub,
BitBucket, and others.\footnote{Open-source variants
(GitLab, Launchpad, Gitorious) of social code platforms allow universities,
labs, or individuals to host the necessary infrastructure on their own
servers.} Besides offering version-controlled source repositories, these
platforms offer many additional features:

\begin{enumerate}

\item Features for managing third-party
contributions (so-called ``pull requests'').

\item Code reviews.

\item References between different project assets (issues, pull
requests, comments), even across projects hosted on the same platform.

\item Social networking: followers, stars, watchers.

\end{enumerate}

By leveraging these and related features, the management of third-party
contributions can be simplified significantly for both contributors and
maintainers. Code reviews enable detailed discussions about proposed changes
and extensions. Continuous integration features enable the (typically cloud-hosted)
infrastructure to run automated builds and tests for each change set
or commit, even before it is incorporated (``merged'') into the mainline
source tree.

As a result, standard open-source culture is such that:

\begin{enumerate}

\item projects can be forked or taken over;

\item discovery and understanding of bug reports is greatly facilitated by
connecting tickets, changesets, branches, etc. through cross references, and
integrating them into open discussions;

\item issues or regressions caused by third-party contributions are often
quickly discovered through continuous integration; moreover, they can be
discussed and addressed conveniently through follow-up commits.

\end{enumerate}

Adopting some of the above-mentioned aspects of mainstream OSS development
could result in a number of benefits for academic research and collaboration:

\begin{enumerate}

\item it is easier to compare with prior work, and to build on it, if that work is
developed in the open, with easily reproducible builds and build scripts;

\item third-party improvements can be incorporated easily upstream, thereby
increasing the value of the original artifact, and its reusability;

\item open, reusable artifacts can be the starting point for bootstrapping on
existing research projects, and even for fruitful research collaborations.

\end{enumerate}

We have two suggestions for achieving these benefits (at least, partially)
through minor adjustments of both paper review processes and AE processes:

First, supplementary material submitted together with research papers could be
allowed to evolve if hosted on a social code platform. Authors should not be
penalized for improving their supplementary material between the time of paper
submission and the time when reviewers chose to look at the material (if
desired). This would encourage authors to continue to improve their
supplementary material in the weeks and months following the initial
submission. Furthermore, it would encourage people to develop artifacts in a
reproducible way from the start, as opposed to quickly throwing together an
artifact in the typically short time between paper notification and artifact
submission. This would also encourage a more disciplined process for building
and developing artifacts.

Second, AE processes could be adjusted in several ways. Clear preference
should be given to artifacts hosted in social code platforms rather than
artifacts provided in the form of virtual machine images. Calls for artifacts
could suggest to authors to host setup instructions and tutorials directly on
their projects' platforms via wikis or other means. Furthermore, if awards for
best artifacts are given out, they should be given only in exceptional cases
to artifacts based on VM images.

Collectively, we believe that these directions have the potential to
contribute positively to both scientific progress and the software community
at large.


% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{bib}

% The bibliography should be embedded for final submission.

\end{document}